\textbf{SPECFEM3D}\\ \textbf{Cartesian}\\ \textbf{User Manual}

### 2024-01-13 doc/USER_MANUAL/13_changing_the_model.tex

To implement your own external tomographic model(s), you must provide
your own external tomography file(s), and choose between two possible
options:\newline
 \indent (1) set in the \texttt{DATA/Par\_file} the model parameter \texttt{MODEL
= tomo}, \newline
 or for more user control: \newline
 \indent (2) set in the \texttt{DATA/Par\_file} the model parameter \texttt{MODEL
= default}, define the negative \texttt{material\_ID} identifier for
each element in the file \texttt{MESH/materials\_file} and use the
following format in the file \texttt{MESH/nummaterial\_velocity\_file}
when using CUBIT to construct your mesh (see also section \ref{subsec:Exporting-the-Mesh}):
\begin{verbatim}
domain_ID material_ID tomography elastic file_name 1
\end{verbatim}
where: \newline
 \indent \texttt{domain\_ID} is 1 for acoustic or 2 for elastic materials,
\newline
 \indent \texttt{material\_ID} a negative, unique identifier (i.e.,
-1,-2,...), \newline
 \indent \texttt{tomography} keyword for tomographic material definition,
\newline
 \indent \texttt{elastic} keyword for elastic material definition,
\newline
 \indent \texttt{file\_name} the name of the tomography file and
1 a positive unique identifier.\newline

@@

\noindent
Each data record line provides the velocity model values in a format like:
{\small
\begin{verbatim}
# data record format: purely isotropic model
x y z   vp    vs    rho
..
\end{verbatim}
}
\noindent
or
{\small
\begin{verbatim}
# data record format: anelastic isotropic model
x y z   vp    vs    rho Qp  Qs
..
\end{verbatim}
}
\noindent
where  \texttt{x}, \texttt{y}, \texttt{z} are the grid point position, \texttt{vp}, \texttt{vs} and \texttt{rho} the P- and S-wave speeds and density, and \texttt{Qp} and \texttt{Qs} the quality factors for P- and S-wave speeds. The quality factors are optional and can be omitted for purely elastic models (the tomography routine will recognize both formats). Internally, the quality factors will be converted to bulk and shear attenuation values,
$Q_{\kappa}$ and $Q_{\mu}$ respectively \citep{AnHa78}.
Note that Qmu is always equal to Qs, but Qkappa is in general not equal to Qp. To convert one to the other see doc/note\_on\_Qkappa\_versus\_Qp.pdf and utils/attenuation/conversion\_from\_Qkappa\_Qmu\_to\_Qp\_Qs\_from\_Dahlen\_Tromp\_959\_960.f90.
For simulations with attenuation, please note that the Vp- and Vs-velocities of your model are given for a reference
frequency. To change this reference frequency, you change the value
of \texttt{ATTENUATION\_f0\_REFERENCE} defined (in Hz) in input file \texttt{DATA/Par\_file}.\newline

The code uses a constant $Q$ quality factor, write(IMAIN,*) "but approximated based on a series of Zener standard linear solids (SLS).
The approximation is thus performed in a given frequency band determined based on that \texttt{ATTENUATION\_f0\_REFERENCE} reference frequency.

@@

The user can implement his own interpolation algorithm for the tomography
model by changing the routine \texttt{model\_tomography.f90} located
in the \texttt{src/generate\_databases/} directory. Moreover, for
models that involve both fully defined materials and a tomography
description, the \texttt{nummaterial\_velocity\_file} has multiple
lines each with the corresponding suitable format described above.

@@

In some cases it may be desirable to have an external tomography model that is described in more than one file. For example, in cases like southern California, the length scale of variation in the structure of the wave speed model is much shorter in the sedimentary basin models within the upper 15~km. Therefore one might want to use an external tomography file that is sampled with, say, $\Delta x = 1000$~m, $\Delta y = 1000$~m, and $\Delta z = 250$~m in the uppermost 15~km, and then use $\Delta x = 2000$~m, $\Delta y = 2000$~m, and $\Delta z = 1000$~m below a depth of 15~km. If these intervals are chosen appropriately, then it will result in a pair of external tomography file that is much smaller than the alternative of having a single file with the fine discretization. In this case  \texttt{nummaterial\_velocity\_file} is
%
\begin{verbatim}
2  -1 tomography elastic file_above_15km.xyz 1
2  -2 tomography elastic file_below_15km.xyz 1
\end{verbatim}
%
and the files \texttt{file\_above\_15km.xyz} and \texttt{file\_below\_15km.xyz} will need to reside in \texttt{DATA/}.  All entries in the second column of \texttt{materials\_file} will need to be either \texttt{-1} or \texttt{-2}, depending on whether the element is above or below 15~km depth. In this sense, the construction of the mesh and the discretization of the external model will generally be done in tandem.\newline

@@

To use your own external model, you can set in the \texttt{DATA/Par\_file}
the model parameter \texttt{MODEL = external}. Three-dimensional acoustic
and/or (an)elastic (attenuation) models may then be superimposed onto
the mesh based upon your subroutine in \texttt{model\_external\_values.f90}
located in directory \texttt{src/generate\_databases}. The call to
this routine would be as follows
\begin{verbatim}
call model_external_values(xmesh, ymesh, zmesh, &
            rho, vp, vs, qkappa_atten, qmu_atten, iflag_aniso, idomain_id)
\end{verbatim}
Input to this routine consists of:
\begin{description}
\item [{\texttt{xmesh,ymesh,zmesh}}] location of mesh point
\end{description}
Output to this routine consists of:
\begin{description}
\item [{\texttt{rho,vp,vs}}] isotropic model parameters for density $\rho$
($kg/m^{3}$), $v_{p}$ ($m/s$) and $v_{s}$ ($m/s$)
\item [{\texttt{qkappa\_atten,qmu\_atten}}] Bulk and shear wave quality factor: $0<Q_{\kappa,\mu}<9000$.
Note that Qmu is always equal to Qs, but Qkappa is in general not equal to Qp. To convert one to the other see doc/note\_on\_Qkappa\_versus\_Qp.pdf and utils/attenuation/conversion\_from\_Qkappa\_Qmu\_to\_Qp\_Qs\_from\_Dahlen\_Tromp\_959\_960.f90.
\item [{\texttt{iflag\_aniso}}] anisotropic model flag, $0$ indicating
no anisotropy or $1$ using anisotropic model parameters as defined
in routine file \texttt{model\_aniso.f90}
\item [{\texttt{idomain\_id}}] domain identifier, $1$ for acoustic, $2$
for elastic, $3$ for poroelastic materials.
\end{description}
Note that the resolution and maximum value of anelastic models are
truncated. This speeds the construction of the standard linear solids
during the meshing stage. To change the resolution, currently at one
significant figure following the decimal, or the maximum value (9000),
consult \texttt{setup/constants.h}.

@@

An example demonstrating the use SEP models is given in:
\texttt{EXAMPLES/meshfem3D\_examples/sep\_bathymetry}.

>>>>

To implement your own external tomographic model(s), you must provide
your own external tomography file(s), and choose between two possible
options:\newline
 \indent (1) set in the \texttt{DATA/Par\_file} the model parameter \texttt{MODEL
= tomo}, \newline
 or for more user control: \newline
 \indent (2) set in the \texttt{DATA/Par\_file} the model parameter \texttt{MODEL
= default}, define the negative \texttt{material\_ID} identifier for
each element in the file \texttt{MESH/materials\_file} and use the
following format in the file \newline
\texttt{MESH/nummaterial\_velocity\_file}
when using CUBIT to construct your mesh (see also section \ref{subsec:Exporting-the-Mesh}):
\begin{verbatim}
domain_ID material_ID tomography elastic file_name 1
\end{verbatim}
where: \newline
 \indent \texttt{domain\_ID} is 1 for acoustic or 2 for elastic materials,
\newline
 \indent \texttt{material\_ID} a negative, unique identifier (i.e.,
-1,-2,...), \newline
 \indent \texttt{tomography} keyword for tomographic material definition,
\newline
 \indent \texttt{elastic} keyword for elastic material definition,
\newline
 \indent \texttt{file\_name} the name of the tomography file and
1 a positive unique identifier.\newline

@@

\noindent
Each data record line provides the velocity model values in a format like:
{\small
\begin{verbatim}
# data record format: purely isotropic model
x y z   vp    vs    rho
..
\end{verbatim}
}
\noindent
or
{\small
\begin{verbatim}
# data record format: anelastic isotropic model
x y z   vp    vs    rho    Qp    Qs
..
\end{verbatim}
}
\noindent
where  \texttt{x}, \texttt{y}, \texttt{z} are the grid point position, \texttt{vp}, \texttt{vs} and \texttt{rho} the P- and S-wave speeds and density, and \texttt{Qp} and \texttt{Qs} the quality factors for P- and S-wave speeds. The quality factors are optional and can be omitted for purely elastic models (the tomography routine will recognize both formats).\newline

\noindent
Internally, the quality factors will be converted to bulk and shear attenuation values,
$Q_{\kappa}$ and $Q_{\mu}$ respectively \citep{AnHa78}.
Note that Qmu is always equal to Qs, but Qkappa is in general not equal to Qp.
To convert one to the other see \texttt{doc/note\_on\_Qkappa\_versus\_Qp.pdf}. There is also a source code file\newline \texttt{conversion\_from\_Qkappa\_Qmu\_to\_Qp\_Qs\_from\_Dahlen\_Tromp\_959\_960.f90}
provided in folder \texttt{utils/small\_utilities/attenuation/}.
For simulations with attenuation, please note that the Vp- and Vs-velocities of your model are given for a reference
frequency. To change this reference frequency, you change the value
of \texttt{ATTENUATION\_f0\_REFERENCE} defined (in Hz) in input file \texttt{DATA/Par\_file}.\newline

\noindent
The code uses a constant $Q$ quality factor, write(IMAIN,*) "but approximated based on a series of Zener standard linear solids (SLS).
The approximation is thus performed in a given frequency band determined based on that \texttt{ATTENUATION\_f0\_REFERENCE} reference frequency.\newline

@@

\noindent
The user can implement his own interpolation algorithm for the tomography
model by changing the routines in file \texttt{model\_tomography.f90} located
in the \texttt{src/generate\_databases/} directory. Moreover, for
models that involve both fully defined materials and a tomography
description, the \texttt{nummaterial\_velocity\_file} has multiple
lines each with the corresponding suitable format described above.\newline

\noindent
Finally, the tomography file can have as many comment lines starting with a hash sign "\#" as you like.
In case you want to provide a tomography file with regular gridded lon/lat/depth coordinates instead of x/y/z,
you can add a comment line like this into the header section of the file:
{\small
\begin{verbatim}
# coordinate format   : lon / lat / depth
\end{verbatim}
}
\noindent
This can be helpful when working with a simulation setup for a specific local region using a corresponding UTM-projection (and topography).
In such cases, it is more convenient to use lon/lat/depth coordinates instead of UTM coordinates directly.\newline
As an example for how to create such a tomography file, a python script \texttt{run\_convert\_IRIS\_EMC\_netCDF\_2\_tomo.py} is provided in
folder \texttt{utils/scripts/} to extract a tomography file from an \href{https://ds.iris.edu/ds/products/emc-earthmodels/}{IRIS EMC model}
that gets provided in netCDF format.

@@

In some cases it may be desirable to have an external tomography model that is described in more than one file. For example, in cases like Southern California, the length scale of variation in the structure of the wave speed model is much shorter in the sedimentary basin models within the upper 15~km. Therefore one might want to use an external tomography file that is sampled with, say, $\Delta x = 1000$~m, $\Delta y = 1000$~m, and $\Delta z = 250$~m in the uppermost 15~km, and then use $\Delta x = 2000$~m, $\Delta y = 2000$~m, and $\Delta z = 1000$~m below a depth of 15~km. If these intervals are chosen appropriately, then it will result in a pair of external tomography file that is much smaller than the alternative of having a single file with the fine discretization. In this case  \texttt{nummaterial\_velocity\_file} is
%
\begin{verbatim}
2  -1 tomography elastic file_above_15km.xyz 1
2  -2 tomography elastic file_below_15km.xyz 1
\end{verbatim}
%
and the files \texttt{file\_above\_15km.xyz} and \texttt{file\_below\_15km.xyz} will need to reside in \texttt{DATA/}.  All entries in the second column of \texttt{materials\_file} will need to be either \texttt{-1} or \texttt{-2}, depending on whether the element is above or below 15~km depth. In this sense, the construction of the mesh and the discretization of the external model will generally be done in tandem.\newline

@@

To use your own external model, you can set in the \texttt{DATA/Par\_file}
the model parameter \texttt{MODEL = external}. Three-dimensional acoustic
and/or (an)elastic (attenuation) models may then be superimposed onto
the mesh based upon your subroutine in \texttt{model\_external\_values.f90}
located in directory \texttt{src/generate\_databases}. The call to
this routine would be as follows
\begin{verbatim}
call model_external_values(xmesh, ymesh, zmesh, &
            rho, vp, vs, qkappa_atten, qmu_atten, iflag_aniso, idomain_id)
\end{verbatim}
Input to this routine consists of:
\begin{description}
\item [{\texttt{xmesh,ymesh,zmesh}}] location of mesh point
\end{description}
Output to this routine consists of:
\begin{description}
\item [{\texttt{rho,vp,vs}}] isotropic model parameters for density $\rho$
($kg/m^{3}$), $v_{p}$ ($m/s$) and $v_{s}$ ($m/s$)
\item [{\texttt{qkappa\_atten,qmu\_atten}}] Bulk and shear wave quality factor: $0<Q_{\kappa,\mu}<9000$.
Note that Qmu is always equal to Qs, but Qkappa is in general not equal to Qp. To convert one to the other see\newline
\texttt{doc/note\_on\_Qkappa\_versus\_Qp.pdf} and source code file\newline
\texttt{conversion\_from\_Qkappa\_Qmu\_to\_Qp\_Qs\_from\_Dahlen\_Tromp\_959\_960.f90}\newline
in folder \texttt{utils/small\_utilities/attenuation}.
\item [{\texttt{iflag\_aniso}}] anisotropic model flag, $0$ indicating
no anisotropy or $1$ using anisotropic model parameters as defined
in routine file \texttt{model\_aniso.f90}
\item [{\texttt{idomain\_id}}] domain identifier, $1$ for acoustic, $2$
for elastic, $3$ for poroelastic materials.
\end{description}
Note that the resolution and maximum value of anelastic models are
truncated. This speeds the construction of the standard linear solids
during the meshing stage. To change the resolution, currently at one
significant figure following the decimal, or the maximum value (9000),
consult \texttt{setup/constants.h}.

@@

An example demonstrating the use of SEP models is given in:\newline
\texttt{EXAMPLES/applications/meshfem3D\_examples/sep\_bathymetry}.\newline

\noindent
Please consider adding improvements and/or new model implementations to this package - feel free to contribute!
